Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-10-10 prop_letter.tex
prop_letter.tex \section{Gleason} Clean up the part moved to Born Sec.

\section{Born}

@@

Recall in Sec.~\ref{sec:fuzzy}, we illustrate the non-constructive
postulates for the quantum probability measures first, and then introduce
the Born rule which induces a quantum probability measures from each
state. Since the postulates for QIVPMs in Sec.~\ref{sec:Interval-Uncertainty}
are also non-constructive, we want to find a simple rule to construct
a QIVPM as well. When we generalized quantum probability measures
to QIVPMs, we changed the range from the unit interval~$\left[0,1\right]$
to a set of intervals~$\mathscr{I}$. This replacement suggests us
to find a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
which can be used to replace a real-valued probability~$\mu\left(P\right)$
by an interval-valued one~$\iota\left(\mu\left(P\right)\right)$.
In other words, compositing $\iota$ with a quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
produces an interval-valued function~$\iota\circ\mu:\events\overset{\mu}{\rightarrow}\left[0,1\right]\overset{\iota}{\rightarrow}\mathscr{I}$
which has the same domain and range as a QIVPM. However, the composited
function~$\iota\circ\mu$ may or may not be a QIVPM. For example,
in a four-dimensional Hilbert space, consider the function 
\begin{equation}
\iota_{1}\left(x\right)=\begin{cases}
\imposs\,, & \textrm{ if }x=0\,;\\
\left[0,\tfrac{1}{2}\right]\,, & \textrm{ if }x\in\left(0,\tfrac{1}{4}\right]\,;\\
\unknown\,, & \textrm{ if }x\in\left(\tfrac{1}{4},\tfrac{3}{4}\right)\,;\\
\left[\tfrac{1}{2},1\right]\,, & \textrm{ if }x\in\left[\tfrac{3}{4},1\right)\,;\\
\necess\,, & \textrm{ if }x=1\,.
\end{cases}\label{eq:iota}
\end{equation}
While the function~$\iota_{1}\circ\mu_{\rho_{0}}^{\mathrm{\mathrm{B}}}$
induced by $\rho_{0}=\frac{1}{4}\cdot\proj{0}+\frac{1}{4}\cdot\proj{1}+\frac{1}{4}\cdot\proj{2}+\frac{1}{4}\cdot\proj{3}$
is a QIVPM, the function~$\iota_{1}\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
induced by $\rho_{1}=\frac{1}{8}\cdot\proj{0}+\frac{1}{8}\cdot\proj{1}+\frac{1}{8}\cdot\proj{2}+\frac{5}{8}\cdot\proj{3}$
is \emph{not} a QIVPM~\citep{HOSTunpublished}. In order not to verify
whether $\iota\circ\mu$ is a QIVPM for each individual case, it is
desirable to have some criteria to discriminate different cases. We
will start from the simpler criteria for the dimension~$d\le3$,
and then the general case.

@@

@@

In Sec.~\ref{sec:Interval-Uncertainty}, we discussed how states
consistent with a QIVPM~$\bar{\mu}$ on the projectors in a subspace
of event~$\events'$. In particular, if $\events'$ commutes, $\bar{\mu}$
restricting on $\events'$ is essentially a classical IVPM, and Thm.~\ref{thm:Shapley}
claims that there is always a state~$\rho$ consistent with $\bar{\mu}$.
In this section, we want to shift our attention to consider states
consistent with $\bar{\mu}$ on \emph{every} event, $\coreBorn\left(\bar{\mu},\events\right)$,
so that these states could characterize $\bar{\mu}$ comprehensively.
When the intervals are infinitely precise, i.e., $\mathscr{I}_{\infty}=\set{\left[x,x\right]}{x\in\left[0,1\right]}$,
a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{\infty}$ becomes
quantum probability measure~\citep{HOSTunpublished}. In this situation,
Gleason's theorem claims that there is always a unique state consistent
with $\bar{\mu}$~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum},
and this state can easily induce $\bar{\mu}$ according to the Born
rule~\citep{Born1983,peres1995quantum,544199,Jaeger2007}. However,
if we consider general QIVPMs mapping to an \emph{arbitrary} set of
intervals, there is a QIVPM that no state can be consistent with it.

@@

In other words, if the measurement intervals are too imprecise, the
results we get might contradict to each other, and exclude any possible
state from being consistent with the measurement results. We will
address this problem with a concrete family of intervals and QIVPMs,
and leave for another time the general framework~\citep{HOSTunpublished}.
Consider the following intervals 
\begin{equation}
\begin{aligned}\mathscr{I}_{n}= & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\\
 & \cup\{\imposs,\necess\}
\end{aligned}
\end{equation}
which are overlapping intervals with length~$1/n$ except for the
intervals with length~$\left(2n\right)^{-1}$ at each end. As shown
in Table~\ref{tab:I_n}, for each interval~$[\ell,r]$, we define
$\textrm{Int}\left([\ell,r]\right)$, the non-overlapping interior
part of $[\ell,r]$.
\begin{table}
\caption{\label{tab:I_n}Intervals in $\mathscr{I}_{n}$, their lengths, and
their interiors, where $2\mid k$ means $k$ is even, and $2\nmid k$
means $k$ is odd. Although intervals~$[\ell,r]$ are overlapping,
their interiors~$\textrm{Int}\left([\ell,r]\right)$ are non-overlapping.}

\begin{tabular}{cccc}
\toprule 
\addlinespace
$k$ & $[\ell,r]$ & $r-\ell$ & $\textrm{Int}\left([\ell,r]\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
 & $\imposs$ & $0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0$ & $\left[0,\tfrac{1}{2n}\right]$ & $\tfrac{1}{2n}$ & $\left(0,\tfrac{0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
$0<k<2n$\tabularnewline
and $2\nmid k$\tabularnewline
\end{tabular} & $\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$ & $\tfrac{1}{n}$ & $\left(\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
$0<k<2n$\tabularnewline
and $2\mid k$\tabularnewline
\end{tabular} & \multirow{1}{*}{$\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$} & $\tfrac{1}{n}$ & $\left[\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$2n$ & $\left[\tfrac{2n-1}{2n},1\right]$ & $\tfrac{1}{2n}$ & $\left[\tfrac{2n-0.5}{2n},1\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
 & $\necess$ & $0$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Since their interior parts are non-overlapping, we can define a function~$\iota_{n}:\left[0,1\right]\rightarrow\mathscr{I}_{n}$
by $\iota_{n}\left(x\right)=[\ell,r]$ if $x\in\textrm{Int}\left([\ell,r]\right)$
to map a real-valued probability to an interval-valued one. This function
can then induce a QIVPM from a state according to the following theorem:

\begin{thm}[A finite-precision Extension of the Born rule]\label{thm:Finite-precision-Born}For
any state $\rho$ on a three-dimensional Hilbert space, a function~$\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}:\events\rightarrow\mathscr{I}_{n}$
defined by $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}\left(P\right)=\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$
is a QIVPM, called a Born QIVPM.\end{thm}

To prove that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$ is a QIVPM,
it is sufficient to check Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
Given $x_{0}$, $x_{1}$, and $x_{0}+x_{1}\in\left[0,1\right]$,
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\label{eq:iota-add-1}
\end{equation}
implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}$ and $P_{1}$ are orthogonal. When $P_{0}$ and $P_{1}$
are not orthogonal on a three-dimensional Hilbert space, we have two
possibilities:
\begin{itemize}
\item Their complements~$\mathbb{1}-P_{0}$ and $\mathbb{1}-P_{1}$ are
orthogonal so that we can apply Eq.~(\ref{eq:iota-add-1}) on their
complements.
\item One of them projects onto a subspace of what the other one projects
onto.
\end{itemize}
\noindent Both cases deduce Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
easily~\citep{HOSTunpublished} so that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
is indeed a QIVPM.

In contrast with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM},
there is always a state consistent with a Born QIVPM $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
because $\rho\in\coreBorn\left(\left[\mu_{\rho}^{\mathrm{B}}\right]_{n},\events\right)$.
Moreover, with respect to a notion of distance between two states~$\rho$
and $\rho'$~\citep{544199}, 
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|\,,\label{eq:norm}
\end{equation}
the collection of states consistent with $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower.

\begin{thm}[A finite-precision extension of weak Gleason's theorem
for Born QIVPMs]\label{thm:Finite-precision-Gleason}In a three-dimensional
Hilbert space, given any state~$\rho$, if a state $\rho'\in\coreBorn\left(\left[\mu_{\rho}^{\mathrm{B}}\right]_{n},\events\right)$,
then their distance $\left\Vert \rho-\rho'\right\Vert \le\frac{1}{n}$.\end{thm}

\noindent Given a state~$\rho'$ consistent with $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$,
we have $\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$
for any 1-dimensional projector~$\proj{\psi}$. Since the maximum
length of the intervals in $\mathscr{I}_{n}$ is $1/n$, it is also
the upper bound of the difference $\left|\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right|=\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$.
 This implies $\left\Vert \rho-\rho'\right\Vert \le\frac{1}{n}$ by
definition.

In the limiting case when the intervals are infinitely precise real
numbers covering the real line, the original Gleason theorem is recovered,
and can be formulated in our framework as follows.

>>>>

\section{Born\label{sec:Born}}

@@

Recall in Sec.~\ref{sec:fuzzy}, we illustrate the non-constructive
postulates for the quantum probability measures first, and then introduce
the Born rule which induces a quantum probability measures from each
state~\citep{Born1983,peres1995quantum,544199,Jaeger2007}. Since
the postulates for QIVPMs in Sec.~\ref{sec:Interval-Uncertainty}
are also non-constructive, we want to find a simple rule to construct
a QIVPM as well. When we generalized quantum probability measures
to QIVPMs, we changed the range from the unit interval~$\left[0,1\right]$
to a set of intervals~$\mathscr{I}$. This replacement suggests us
to find a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
which can be used to replace a real-valued probability~$\mu\left(P\right)$
by an interval-valued one~$\iota\left(\mu\left(P\right)\right)$.
In other words, compositing $\iota$ with a quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
produces an interval-valued function~$\iota\circ\mu:\events\overset{\mu}{\rightarrow}\left[0,1\right]\overset{\iota}{\rightarrow}\mathscr{I}$
which has the same domain and range as a QIVPM. However, the composited
function~$\iota\circ\mu$ may or may not be a QIVPM. For example,
in a four-dimensional Hilbert space, consider the function 
\begin{equation}
\iota_{1}\left(x\right)=\begin{cases}
\imposs\,, & \textrm{ if }x=0\,;\\
\left[0,\tfrac{1}{2}\right]\,, & \textrm{ if }x\in\left(0,\tfrac{1}{4}\right]\,;\\
\unknown\,, & \textrm{ if }x\in\left(\tfrac{1}{4},\tfrac{3}{4}\right)\,;\\
\left[\tfrac{1}{2},1\right]\,, & \textrm{ if }x\in\left[\tfrac{3}{4},1\right)\,;\\
\necess\,, & \textrm{ if }x=1\,.
\end{cases}\label{eq:iota}
\end{equation}
While the function~$\iota_{1}\circ\mu_{\rho_{0}}^{\mathrm{\mathrm{B}}}$
induced by $\rho_{0}=\frac{1}{4}\cdot\proj{0}+\frac{1}{4}\cdot\proj{1}+\frac{1}{4}\cdot\proj{2}+\frac{1}{4}\cdot\proj{3}$
is a QIVPM, the function~$\iota_{1}\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
induced by $\rho_{1}=\frac{1}{8}\cdot\proj{0}+\frac{1}{8}\cdot\proj{1}+\frac{1}{8}\cdot\proj{2}+\frac{5}{8}\cdot\proj{3}$
is \emph{not} a QIVPM~\citep{HOSTunpublished}. In order not to verify
whether $\iota\circ\mu$ is a QIVPM for each individual case, it is
desirable to have some criteria to discriminate different cases. We
will start from the simpler criteria for the dimension~$d\le3$,
and then the general case.

@@

\yutsung{The other idea: We never proved or find an counterexample
that a non-convex QIVPM has empty core, and its expectation value
is not the same as defined from core... If it is true, the convex
condition may not be necessary, but this might be hard to prove...}

@@

In Sec.~\ref{sec:Interval-Uncertainty}, we discussed how states
consistent with a QIVPM~$\bar{\mu}$ on the projectors in a subspace
of event~$\events'$. In particular, if $\events'$ commutes, $\bar{\mu}$
restricting on $\events'$ is essentially a classical IVPM, and Thm.~\ref{thm:Shapley}
claims that there is always a state~$\rho$ consistent with $\bar{\mu}$.
In this section, we want to shift our attention to consider states
consistent with $\bar{\mu}$ on \emph{every} event, $\coreBorn\left(\bar{\mu},\events\right)$.
When the intervals are infinitely precise, i.e., $\mathscr{I}_{\infty}=\set{\left[x,x\right]}{x\in\left[0,1\right]}$,
a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{\infty}$ becomes
quantum probability measure~\citep{HOSTunpublished}. In this situation,
Gleason's theorem claims that there is always a unique state consistent
with $\bar{\mu}$~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.

\noindent However, if we consider general QIVPMs mapping to an \emph{arbitrary}
set of intervals, there is a QIVPM that no state can be consistent
with it.

@@

In other words, if the measurement intervals are too imprecise, the
results we get might contradict to each other, and exclude any possible
state from being consistent with the measurement results. In contrast
with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM}, if a QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{B}}$
is induced by a conventional quantum probability measure~$\mu_{\rho}^{\mathrm{B}}$
and a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$ defined
in Sec.~\ref{sec:Born}, there is always a state consistent with
it because $\rho\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$.
Moreover, with respect to a notion of distance between two states~$\rho$
and $\rho'$~\citep{544199}, 
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|\,,\label{eq:norm}
\end{equation}
the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower.

\begin{thm}[A finite-precision extension of weak Gleason's theorem
for Born QIVPMs]\label{thm:Finite-precision-Gleason}Consider a three-dimensional
Hilbert space, and an interval-valued function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
with the maximum length of intervals in $\mathscr{I}$ is $\delta$.
Given any state~$\rho$, if a state $\rho'\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$,
then their distance $\left\Vert \rho-\rho'\right\Vert \le\delta$.\end{thm}

\noindent Given a state~$\rho'$ consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$,
we have $\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota\left(\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$
for any 1-dimensional projector~$\proj{\psi}$. Since the maximum
length of the intervals in $\mathscr{I}$ is $\delta$, it is also
the upper bound of the difference $\left|\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right|=\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$.
This implies $\left\Vert \rho-\rho'\right\Vert \le\delta$ by definition.
